---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
===================================
##Part1

After you've downloaded the zip file and unzipped it, you should have "activity.csv" in your working directory.  
Let's start off by loading the data from "activity.csv" and take a look at it.
```{r, echo=TRUE}
activity<-read.csv("activity.csv")
summary(activity)
```
 
For the our purpose here, let's get rid of the NAs.
```{r, echo=TRUE}
complete<-complete.cases(activity)
act_comp<-activity[complete,]
summary(act_comp)
```
Now let's calculate the total steps taken each day.  
Let's also omit missing values here.
```{r, echo=TRUE}
sums<-tapply(act_comp$steps,act_comp$date,sum)
summary(sums)
```
For dates that only have 0 steps, the sum value for those dates gets calculated as NA. Let's change that to 0 instead.
```{r}
sums[is.na(sums)]<-0
summary(sums)
```

###Histogram for Part 1
So now that we've cleaned that up a bit, let's create a histogram for the total steps taken per day
```{r, echo=TRUE}
library(ggplot2)

sums_dat<-as.data.frame(sums)

g<-ggplot(sums_dat,aes(sums))
g<-g + geom_histogram()
g<-g + labs(x = "Total Steps")
g
```

###Mean and Median for Part 1
Let's see what the mean is of the total steps taken per day.
```{r}
mean(sums)
```
Let's also see the median.
```{r}
median(sums)
```
  
  

##Part 2

###Time Series Plot for Part 2
Let's plot average steps for each interval across all dates against the intervals.  
For the purpose of the plot, let's again use the act_comp data frame created in Part 1 which only includes complete cases.

```{r}
df<-aggregate(data=act_comp,steps~interval,mean)

gg<-ggplot(data=df, aes(x=interval,y=steps))
gg<-gg + geom_line(colour="blue")
gg<-gg + labs(x="Interval", y="Average Number of Steps")
gg


```

##Highest Average Number of Steps for Part 2
So which interval has the highest average number of steps?
```{r}
df[which(df$steps==
             max(df$steps)),
   ]$interval
```
  
  

##Part 3
So how many rows have missing values?
```{r}
summary(activity)
```
2304 rows according to the summary  

###Imputing the Data for Part 3
**Note that this is probably not going to be the most concise way of doing so  
So now let's impute the NAs. For our purpose here, I'll use the means based on the intervals.  
I'll start off by calculating the means and then merging it to the data frame as a new column. This new column appears as "steps.y"
```{r}
means<-aggregate(activity$steps,list(activity$interval),mean, na.rm=T)
colnames(means)<-c("interval","steps")
merged_dat<-merge(activity, means, by="interval")
summary(merged_dat)
```
Now let's subset for just the rows with NAs.
```{r}
dat.na<-merged_dat[is.na(merged_dat$steps.x),]
```
Let's now cut the NAs out and create a data frame with the dates, intervals, and means
```{r}
library(dplyr)
dat.na.2<-select(dat.na, -steps.x)
dat.na.2<-rename(dat.na.2,steps=steps.y)
dat.na.2<-select(dat.na.2, steps, date, interval)
head(dat.na.2)
```
And now, let's combine this data frame to the act_comp data frame previously created (again, this is just the original data frame with complete cases)
```{r}
act.2<-rbind(act_comp,dat.na.2)
summary(act.2)
```

###Histogram for Part 3
Let's now see what the total steps per day looks like.
```{r}
sums.2<-as.data.frame(tapply(act.2$steps,act.2$date,sum))
colnames(sums.2)<-"sums"

library(ggplot2)
g<-ggplot(sums.2,aes(sums))
g<-g + geom_histogram()
g<-g + labs(x = "Total Steps")
g
```

What does the mean look like now?
```{r}
mean(sums.2$sums)
```

What does the median look like now?
```{r}
median(sums.2$sums)
```
With the imputed data, both the mean and median are noticeable higher than when NAs were simply dropped
  
  

##Part 4
Let's add a new column to the data frame that shows whether the date is during the week or weekend.  
They should be factor variables
```{r}
library(dplyr)
act.2.days<-mutate(act.2,
                   day=weekdays(as.POSIXct(activity$date)))

act.2.days$day[which(act.2.days$day=="Monday")]<-"weekday"
act.2.days$day[which(act.2.days$day=="Tuesday")]<-"weekday"
act.2.days$day[which(act.2.days$day=="Wednesday")]<-"weekday"
act.2.days$day[which(act.2.days$day=="Thursday")]<-"weekday"
act.2.days$day[which(act.2.days$day=="Friday")]<-"weekday"
act.2.days$day[which(act.2.days$day=="Saturday")]<-"weekend"
act.2.days$day[which(act.2.days$day=="Sunday")]<-"weekend"

act.2.days$day<-as.factor(act.2.days$day)

summary(act.2.days)
```

So how do the number of steps differ from the weekdays to the weekends?  
Let's first calculate the average number of step per interval accross all days, but let's calculate these separately for weekdays and weekends.
```{r}
act.2.wd<-act.2.days[which(act.2.days$day=="weekday"),]
act.2.we<-act.2.days[which(act.2.days$day=="weekend"),]

df2.wd<-as.data.frame(aggregate(data=act.2.wd,steps~interval,mean))
df2.wd<-mutate(df2.wd,day="weekday")
df2.wd$day<-as.factor(df2.wd$day)

df2.we<-as.data.frame(aggregate(data=act.2.we,steps~interval,mean))
df2.we<-mutate(df2.we,day="weekend")
df2.we$day<-as.factor(df2.we$day)
```

###Panel Plot for Part 4
Now that we've calculated the averages, let's compare them on a panel plot

```{r}
df2<-rbind(df2.wd,df2.we)
##So that ggplot can create the two panels we need to combine the weekday and weekend means into one data frame.

library(ggplot2)

ggg<-ggplot(data=df2, aes(x=interval,y=steps))
ggg<-ggg + geom_line(colour="blue")
ggg<-ggg + labs(x="Interval", y="Average Number of Steps")
ggg<-ggg + facet_grid(day~.)
ggg
```